---
title: "CV3D: Analysing Compound-Eye Geometry in R"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{CV3D: Analysing Compound-Eye Geometry in R}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE,
  warning = FALSE
)
```

## Introduction

CV3D provides the R-based quantitative components of a workflow for extracting and analysing compound-eye surface geometry from three-dimensional data. The package can be used directly in R; preprocessing, interactive inspection, parameter selection, and orchestration of the complete workflow are provided by the CV3D UI.

The examples below use reduced subsets of an adult *Drosophila melanogaster* compound eye derived from the micro-computed tomography data described by Schoborg et al. (2019):

> Schoborg TA, Smith SL, Smith LN, Morris HD, Rusan NM (2019). Micro-computed tomography as a platform for exploring Drosophila development. *Development* 146(23): dev176685. doi:10.1242/dev.176685.

```{r load-package}
library(CV3D)
```

All spatial coordinates in CV3D are currently assumed to be expressed in micrometres (µm). The package does not infer or convert spatial units from arbitrary input meshes.

## Example datasets

CV3D contains three small example datasets representing successive stages of the R workflow.

```{r example-data}
data(cv3d_example_surface)
data(cv3d_example_thresholded)
data(cv3d_example_facets)

head(cv3d_example_surface)
head(cv3d_example_thresholded)
head(cv3d_example_facets)
```

`cv3d_example_surface` contains triangle centres and triangle surface normals from a corneal surface mesh. `cv3d_example_thresholded` contains points retained after local-height thresholding. `cv3d_example_facets` contains checked facet-centre coordinates.

## Local surface heights

Local surface height describes the signed displacement of each mesh triangle centre from a reference plane estimated from its surrounding surface. Raised facet centres therefore produce local-height structure that can be used for facet detection.

```{r local-heights}
local_heights <- calculate_local_heights(
  cv3d_example_surface,
  neighbourhood_radius = 5.5,
  cores = 1,
  verbose = FALSE
)

head(local_heights[, c("ID", "x", "y", "z", "local_height")])
```

The neighbourhood value above is appropriate for this reduced example and should not be interpreted as a universal parameter recommendation. In a normal CV3D analysis it is chosen relative to the approximate facet size of the dataset. The scientifically appropriate neighbourhood scale remains dataset-dependent and should be large enough to represent the local eye surface rather than individual facet relief.

`calculate_local_heights()` also returns `local_height_contrast`, a 0--1 peak-enhancing variable. It preserves the historical `10^local_height` transform, clips the transformed values to configurable quantiles, and rescales them to 0--1 for intuitive threshold selection. The raw `local_height` remains the geometric quantity in µm.

Local-height normalization is optional. It rescales values relative to overlapping local neighbourhoods and can reduce spatial variation in local-height magnitude across an eye surface.

```{r normalize-heights}
local_heights_normalized <- normalize_local_heights(
  local_heights,
  neighbourhood_radius = 14.7,
  cores = 1,
  plot_results = FALSE,
  verbose = FALSE
)

head(local_heights_normalized[, c("ID", "local_height", "local_height_norm")])
```

## Facet-candidate condensation

Threshold selection is normally performed interactively in the CV3D UI. The bundled thresholded example therefore starts immediately after that step. `find_facet_candidates_condensed()` moves thresholded points towards local height-weighted modes and condenses nearby converged points into facet candidates.

```{r candidates}
candidates <- find_facet_candidates_condensed(
  cv3d_example_thresholded,
  neighbour_radius = 7.5,
  merge_radius = 4.5,
  cores = 1,
  verbose = FALSE
)

head(candidates)
```

Returned facet-candidate coordinates always correspond to actual input points rather than interpolated centroids.

## Edge detection and facet-neighbour selection

Neighbour selection is a distinct geometric step in the current CV3D workflow. The current UI uses the edge-aware method described here and stores the resulting neighbour graph before optical calculations. The older `find_neighbours()` function remains available for backward compatibility and method comparison, but it is not the preferred neighbour-selection path in the current UI workflow.

### Detecting the sampled eye boundary

`detect_facet_edges()` identifies likely boundary facets from local angular coverage. Around each focal facet, nearby facet centres are used to estimate a local tangent plane. The six nearest candidate directions are projected into that plane and sorted by polar angle. The largest circular angular gap, including the wrap-around gap, is recorded as `edge_angular_gap_deg`.

An interior facet with a complete first-ring neighbourhood normally has candidate directions distributed around the full circle. A boundary facet has a conspicuous unsampled sector. A facet is therefore classified as an edge when its largest angular gap is greater than a user-supplied threshold.

```{r edge-detection}
edge_qc <- detect_facet_edges(
  cv3d_example_facets,
  gap_threshold_deg = 90
)

head(edge_qc)
table(edge_qc$is_edge_facet)
```

The threshold is deliberately explicit rather than fitted automatically. The CV3D UI compares 80, 85, 90, 95, 100, and 105 degrees in a face-on QC figure and asks the user to choose the threshold appropriate for the sampled eye boundary.

### Constructing the edge-aware neighbour graph

`find_neighbours_edge_aware()` combines the boundary classification with a conservative first-ring graph:

1. candidate links are restricted to reciprocal six-nearest-neighbour membership;
2. local core spacing is the median distance to the three closest facet centres;
3. interior-interior links are retained up to 1.5 times the larger endpoint core spacing;
4. links touching a detected edge facet use a stricter factor of 1.4;
5. at detected edge facets only, a near-collinear farther neighbour can be removed by the angular-shadow test when it is at least 1.15 times farther away than the nearer competing neighbour; and
6. no refill step forces edge facets back to a predetermined neighbour count.

The angular-shadow cutoff is calibrated from the retained six-neighbour interior lattice: two-thirds of its median adjacent angular spacing, constrained to 30--45 degrees. Links are removed reciprocally and the focal facet is not pruned below the configured minimum.

```{r neighbours}
facets <- find_neighbours_edge_aware(
  cv3d_example_facets,
  edge_gap_threshold_deg = 90,
  verbose = FALSE
)

head(facets[, c(
  "ID", "neighbours", "number_of_neighbours",
  "is_edge_facet", "edge_angular_gap_deg"
)])
```

The output preserves the selected edge threshold, local core spacing, and per-facet shadow-pruning count. Additional attributes record the derived shadow-angle cutoff, removed shadow links, and the method identifier. These fields are useful for QC and provenance rather than as optical measurements themselves.

## Facet size

Facet size is estimated from the retained neighbour graph. For each facet, CV3D calculates the arithmetic mean of the three-dimensional Euclidean centre-to-centre distances to all assigned neighbours. This is a local spacing proxy for facet diameter, not a direct segmentation of the facet boundary.

A smoothed estimate is additionally calculated by giving equal weight to the focal facet-size estimate and the mean estimate of its neighbouring facets.

```{r facet-size}
facet_sizes <- calculate_facet_size(facets)
head(facet_sizes)

facets <- merge(
  facets,
  facet_sizes,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)
```

All size and distance outputs are in µm because the input coordinates are assumed to be in µm.

## Facet surface normals

CV3D provides two facet-normal estimators.

### Regularised facet-centre envelope

`get_facet_normals_envelope()` is the current CV3D UI default. It reconstructs a triangular envelope directly from the facet-centre point cloud and is intentionally independent of the optical neighbour graph.

The envelope is built from reciprocal six-nearest-neighbour links. Triangular three-cliques form candidate faces, and a candidate face is rejected if any edge is longer than 1.4 times the larger local six-neighbour spacing at its endpoints. The retained envelope is regularised by five Taubin-style smoothing cycles with lambda = 0.30 and mu = -0.31.

For each original facet centre, envelope-face normals are combined with weights proportional to face area times a Gaussian distance term. The Gaussian scale is the local facet spacing multiplied by `envelope_factor`; support is truncated at 2.5 times that scale. The weighted normal is normalised and oriented outward relative to the arithmetic centroid of the facet-centre cloud. No post-neighbour normal averaging is applied.

```{r envelope-normals}
facet_normals <- get_facet_normals_envelope(
  cv3d_example_facets,
  envelope_factor = 1.25,
  verbose = FALSE
)

head(facet_normals[, c("ID", "norm.x", "norm.y", "norm.z")])
```

The current UI offers envelope factors 1, 1.25, 1.5, and 2, with 1.25 as the working default. The 1.25 value is a practical current default informed by CV3D degradation testing, not a claim of universal optimality across all eye shapes or sampling qualities. Smaller values use a more local surface patch; larger values smooth over a broader patch.

### Original CV3D normal estimator

`get_facet_normals()` remains available as the original estimator. It forms triangles from the focal facet and pairs of assigned neighbours, averages the outward local triangle normals, and then averages the focal normal with valid neighbouring normals before renormalisation.

```{r original-normals}
original_normals <- get_facet_normals(
  facets,
  cores = 1,
  plot_results = FALSE,
  verbose = FALSE
)

head(original_normals)
```

The original and envelope estimators therefore differ in both their surface representation and their smoothing strategy. They should not be treated as interchangeable implementations of the same numerical algorithm.

For the remainder of this vignette, the envelope normals are merged into the facet table.

```{r merge-normals}
facets <- merge(
  facets,
  facet_normals,
  by = "ID",
  all.x = TRUE,
  sort = FALSE
)
```

## Standardised face-on plotting

`view_eye_face_on()` provides the standard orientation used by current CV3D QC plots. When no viewing direction is supplied, the mean of the valid unit facet normals defines the outward viewing direction. A screen basis is then constructed from the supplied up vector projected perpendicular to that direction. If requested, the greater face-on extent of the eye is placed vertically.

The same orientation can be reused across plotting layers or between 2D and 3D QC views. In 2D, the function returns the projected coordinates. In 3D, it also returns the `rgl` camera matrix used to reproduce the same face-on orientation interactively.

```{r face-on, fig.width=4.5, fig.height=4.5}
face_on <- view_eye_face_on(
  facets,
  projection = "2D",
  axes = FALSE,
  pch = 16,
  cex = 0.5
)

head(face_on$projected_coordinates)
```

This operation changes only the display orientation. It does not rotate or overwrite the analysed coordinates in the input table.

## Local optical properties

Local inter-facet angles are derived from the mean angular separation between a facet surface normal and its immediate neighbouring normals. Together with facet size, they are used to calculate Snyder's eye parameter and ommatidial sampling frequency. CV3D supports square and hexagonal sampling lattices; the default is hexagonal.

```{r optics}
optics <- get_optic_properties(
  facets,
  lattice = "hexagonal",
  cores = 1,
  plot_results = FALSE,
  verbose = FALSE
)

head(optics)
```

For a square lattice, Snyder's sampling frequency is $1/(2\Delta\phi)$ and the eye parameter is $D\Delta\phi$. For a hexagonal lattice, the corresponding expressions are $1/(\sqrt{3}\Delta\phi)$ and $D\sqrt{3}\Delta\phi/2$. Here, $D$ is the smoothed neighbour-centre spacing used as a proxy for facet diameter (in µm), and $\Delta\phi$ is the local inter-facet angle in radians. The eye parameter is therefore in µm·rad.

`acuity_cpd` is calculated separately as $1/(2\Delta\phi)$ with $\Delta\phi$ in degrees, using the interommatidial-angle relationship used by Feller et al. (2020). CV3D applies this locally to each facet using its mean immediate-neighbour inter-facet angle, so the result is a local anatomical acuity estimate rather than a species-level behavioural acuity measurement. It is independent of the square/hexagonal lattice option.

## Landmark-based coordinate alignment

`align_pointcloud()` places a point cloud into a standardised anatomical coordinate system. The anterior landmark becomes the origin, positive x points from right to left, positive y points from anterior to posterior, and z completes a right-handed coordinate system.

```{r alignment}
landmarks <- data.frame(
  ID = c("anterior", "posterior", "left", "right", "facet_1"),
  x = c(0, 0, 1, -1, 0),
  y = c(0, 2, 1, 1, 1),
  z = c(0, 0, 0, 0, 1),
  norm.x = c(NA, NA, NA, NA, 0),
  norm.y = c(NA, NA, NA, NA, 0),
  norm.z = c(NA, NA, NA, NA, 1)
)

aligned <- align_pointcloud(landmarks)
aligned[, c("ID", "x", "y", "z")]
```

## Projection of viewing directions

Facet viewing directions can be projected from facet centres onto a sphere using `ray_sphere_intersection()`.

```{r ray-sphere}
intersection <- ray_sphere_intersection(
  point = c(0, 0, 0),
  direction = c(1, 0, 0),
  sphere_center = c(0, 0, 0),
  sphere_radius = 10
)

intersection
```

Cartesian projection coordinates can then be expressed as elevation and azimuth with `cartesian_to_view_angles()`.

```{r view-angles}
view_angles <- cartesian_to_view_angles(
  x = c(0, 1, 0, -1, 0),
  y = c(-1, 0, 1, 0, 0),
  z = c(0, 0, 0, 0, 1)
)

view_angles
```

Under the CV3D viewing convention, anterior is 0 degrees azimuth, left is +90 degrees, right is -90 degrees, posterior is -180/180 degrees, and positive z gives positive elevation.

## Complete CV3D workflow

The functions shown here constitute the R-based analysis stages of CV3D. In the current UI workflow, checked Step-04 facet positions are followed by a dedicated Step 04B neighbour-selection stage. The user compares candidate edge-gap thresholds, chooses the boundary threshold, and stores the resulting edge-aware neighbour graph. Step 05A then consumes that stored graph, calculates facet size, estimates normals using the selected normal method, and derives local optical properties.

The complete workflow from 3D image data through surface extraction, facet checking, neighbour QC, landmarking, analysis, face-on QC plotting, and export is guided by the CV3D UI and its accompanying tutorial.
